Robust Estimation, Prediction and Control with \\ Linear Dynamics and Arbitrary Costs

### 2020-01-20 main.tex
Little bit of related work

We leverage known techniques from the uncertain system simulation literature to efficiently compute such a predictor.

@@

For LQ systems, where parameters A,B are unknown and no structure is available.
\begin{itemize}
    \item OFU: \citep{abbasi-yadkori11a}. Summary: $\theta=(A~B)$, ellipsoid in Frobenius norm + controllability constraint, choose and stabilize (Ricatti) an optimistic candidate, cumulative regret in $O(\sqrt{T}\log\frac{1}{\delta})$.
    \item PSRL with TS: \citep{abeille18a}. Summary: RLS estimate + normal perturbation in the ellipsoid + rejection sampling over a controllability constraint, with frequent update, cumulative regret in $\tilde{O}(\sqrt{T})$
    \item Offline estimation and control synthesis: \citep{Dean2017}. Summary: Estimation procedure of $N$ episodes of duration $T$ with gaussian excitation, ellipsoid in L2 norm, robust controller with System Level Synthesis, simple regret in $O(\sqrt{\frac{\log(1/\delta)}{N}})$
    \item \citep{Dean2018}
\end{itemize}

Differences with our work:

@@

    Thus, we do not wish to explore actively the more promising regions as in OFU, nor the more likely region as in TS, nor act randomly as in Coarse-ID.

    We have linear dynamics but our cost is not quadratic. Thus, we can accurately represent planning tasks with discrete/branching decisions involving discontinuities (e.g. collision vs non-collision states).

@@

Robust optimization has been studied in the context of finite Markov Decision Processes (MDP) with uncertain parameters by Iyengar (2005), Nilim and El Ghaoui (2005) and Wiesemann et al. (2013).They show that the main results of Dynamic Programming can be extended to their robust counterparts only when the dynamics ambiguity set verifies certain rectangularity properties, which can be seen as an independance assumption:

@@

We also assume that some bound $S$ over $\|\theta\|_\infty$ is known.
\end{assumption}

@@

Let $\{\Phi_n\}_{n=1}^\infty$ be an $\Real^{p\times d}$-valued stochastic process such that $\Phi_n$ is $F_n$-measurable. Assume that $G$ is a $d\times d$ positive definite matrix. For any $n\geq 0$, define:
\begin{equation*}
    \overline{G}_n = G + \sum_{s=1}^n \Phi_s^\transp \Sigma_p^{-1} \Phi_s \in \Real^{d\times d} \quad S_n = \sum_{s=1}^n \Phi_s^\transp\Sigma_p^{-1}\eta_s \in \Real^{d}.
\end{equation*}
Then, for any $\delta>0$, with probability at least $1-\delta$, for all $n\geq0$,
\begin{align*}
\| S_n \|_{\overline{G}_n^{-1}} \geq \sqrt{2\ln \left(\frac{\det\left(\overline{G}_n\right)^{1/2}}{\delta\det(G)^{1/2}}\right)}.
\end{align*}
\end{proposition}

\begin{theorem}[Confidence ellipsoid]
\label{thm:confidence_ellipsoid}
Under \autoref{assumpt:gaussian-noise}, it holds with probability at least $1-\delta$ that
\begin{align}
    \label{eq:confidence-ellipsoid}
    \| \theta_{N,\lambda}  - \theta\|_{G_{N,\lambda}} \leq \beta_N(\delta),
\end{align}
with
\begin{equation}
    \beta_N(\delta)\eqdef \sqrt{2\ln \left(\frac{\det(G_{N,\lambda})^{1/2}}{\delta\det(\lambda I_d)^{1/2}}\right)}
     + (\lambda d)^{1/2}S.
\end{equation}
\end{theorem}

@@

Another strategy presented in \autoref{sec:tight-polytope} produces a much tighter polytope, at the price of an increased computational cost required by the diagonalization of $G_{N,\lambda}$.

@@

However, \citet{leurent2019interval} showed that this predictor can have very unstable dynamics, even for stable systems, which can cause a rapid build-up of uncertainty and produce over-cautious behaviours. They propose an enhanced predictor which produces tighter and more stable predictions, and whose stability can be assessed by solving a linear matrix inequality, at the price of an additional requirement:

\begin{assumption}
\label{assumpt:metzler}
There exists an orthogonal matrix $P\in\Real^{d\times d}$ such that $P^\transp A_0 P$ is Metzler, that is, all non-diagonal coefficients are non-negative.
\end{assumption}
In practice, this assumption is often verified. It is for instance the case whenever $A_0$ is diagonalizable. The similarity transformation of \citep{Efimov_a2013} provides a method to compute such $P$ whenever there exist $C_1,C_2$ which make $A(\theta)$ and $A_0$ is observable. Note that similar requirements involving the controllability of $A_0$ exist the literature of the LQ setting, see e.g. \citep{abbasi-yadkori11a,abeille18a}.
Under this assumption, without loss of generality up to a change of basis $x\rightarrow P^{-1}x$ that we do not write for the sake of readability, this assumption is equivalent to considering that $A_0$ itself is Metzler.

@@

\subsection{Proof of \autoref{prop:concentration}}

\begin{proof}

Let 
\begin{equation*}
    G_t = \sum_{s=1}^t \Phi_s^\transp \Sigma_p^{-1} \Phi_s \in \Real^{d\times d}
\end{equation*}
And for any $z\in\Real^d$,
\begin{equation*}
    M_t^z = \exp{\left(\inp{z}{S_t} - \frac{1}{2}\|z\|_{G_t}\right)}
\end{equation*}
\begin{equation*}
    D_t^z = \exp{\left(\inp{\Phi_t z}{\eta_t}_{\Sigma_p^{-1}} - \frac{1}{2}\|\Phi_t z\|_{\Sigma_p^{-1}}\right)}
\end{equation*}
Then,
\begin{align*}
    M_t^z &= \exp{\left(\sum_{s=1}^t z^\transp \Phi_s^\transp \Sigma_p^{-1} \eta_s - \frac{1}{2} (\Phi_s z)^\transp\Sigma_p^{-1}(\Phi_s z) \right)} \\
    &= \prod_{s=1}^{t} D_s^z
\end{align*}
and using the Sub-Gaussianity of $\eta_t$
\begin{align*}
    \expectedvalue\left[D_t^z \condbar F_{t-1}\right] = {}& \exp{\left(- \frac{1}{2}\|\Phi_t z\|_{\Sigma_p^{-1}}\right)}\\ &\expectedvalue\left[\exp{\left(\inp{\Phi_t z}{\eta_t}_{\Sigma_p^{-1}}\right)} \condbar F_{t-1}\right]  \\
    \leq {} & \exp{\left(- \frac{1}{2}\|\Phi_t z\|_{\Sigma_p^{-1}}\right)}\\
    &\exp{\left((z^\transp \Phi_t^\transp \Sigma_p^{-1})\Sigma_p(\Sigma_p^{-1} \Phi_t z)\right)}\\
    &= 1
\end{align*}
\begin{align*}
    \expectedvalue\left[M_t^z \condbar F_{t-1}\right] = \left(\prod_{s=1}^{t-1} D_s^z\right) \expectedvalue\left[D_t^z \condbar F_{t-1}\right] \leq M_{t-1}^z
\end{align*}
Showing that $(M_t^z)_{t=1}^\infty$ is indeed a supermartingale and in fact $\expectedvalue[M_t^z]\leq 1$.
It then follows by Doob's upcrossing lemma for supermartingale that $M_\infty^z = \lim_{t\to\infty} M_t^z$ is almost surely well-defined, and so is $M_\tau^z$ for any random stopping time $\tau$.

@@

% The following uses the method of mixtures.
% Let $\Gamma$ be a Gaussian random variable which is independent of all other random variables and whose covariance is $\Sigma^{-1}$. (in our estimator, $\Sigma = \lambda$).

% Define 
% \[
% M_t = \expectedvalue \left[M_t^\Gamma \condbar F_\infty\right]
% \]

% where $F_\infty$ is the tail $\sigma$-algebra of the filtration. Clearly, we still have $\expectedvalue[M_\tau] = \expectedvalue[\expectedvalue[M_\tau^\Gamma|\Gamma]]\leq 1$.

% Let us calculate $M_t$. Let $f$ denote the density of $\Gamma$ and for a positive definite matrix $P$ let $c(P) = \sqrt{(2\pi)^d/\det(P)} = \int \exp (-\frac{1}{2}x^\transp P x)$

@@

\begin{proof}

Hence, for all $x\in\Real^d$,
\begin{align*}
    x^\transp\theta_{N,\lambda}  -x^\transp\theta &= x^\transp G_{N, \lambda}^{-1}\sum_{n=1}^N \Phi_n^\transp \Sigma_p^{-1}\eta_n
    - \lambda x^\transp G_{N, \lambda}^{-1}\theta\\
    &= \inp{x}{\sum_{n=1}^N \Phi_n^\transp \Sigma_p^{-1}\eta_n}_{G_{N, \lambda}^{-1}} - \lambda\inp{x}{\theta}_{G_{N, \lambda}^{-1}}
\end{align*}

>>>>

We leverage recent results from the uncertain system simulation literature to efficiently compute such a predictor.

@@

\paragraph{LQ systems} There is a rich body of literature dedicated to the study of LQ systems, and we focus on those providing non-asymptotic results when the system parameters $A, B$ are unknown. In their seminal work \citet{abbasi-yadkori11a}, consider the problem of cumulative regret minimisation. They follow the \emph{Optimism in the Face of Uncertainty} paradigm from the multi-armed bandit literature, that consists in selecting an optimistic system within a high-confidence region while enforcing a controllability constraint, and computing the corresponding optimal control in closed-form by solving a Riccati equation. They show that this procedures achieves a $\tilde{\cO}\left(T^{1/2}\right)$ cumulative regret. \citet{abeille18a} provide the same result by applying Thomson Sampling rather of optimism to select a dynamical model, and rejection sampling to enforce the controllability constraint.
While previous methods suffered from limited applicability and computationally intractable subroutines, \citep{Dean2018} provide a polynomial-time algorithm which achieves a suboptimal cumulative regret of $\tilde{\cO}\left(T^{2/3}\right)$.

Offline estimation and control synthesis: \citep{Dean2017}. Summary: Estimation procedure of $N$ episodes of duration $T$ with gaussian excitation, ellipsoid in L2 norm, robust controller with System Level Synthesis, simple regret in $O(\sqrt{\frac{\log(1/\delta)}{N}})$

@@

Thus, we do not wish to explore actively the more promising regions as in OFU, nor the more likely region as in TS, nor act randomly as in Coarse-ID They inject exploration noise.

We have linear dynamics but our cost is not quadratic. Quadratic cost: stabilize a system around a reference trajectory known a priori. Thus, we can accurately represent planning tasks with discrete/branching decisions involving discontinuities (e.g. collision vs non-collision states).

@@

\paragraph{Robust planning}
Robust optimization has been studied in the context of finite Markov Decision Processes with uncertain parameters by Iyengar (2005), Nilim and El Ghaoui (2005) and Wiesemann et al. (2013). They show that the main results of Dynamic Programming can be extended to their robust counterparts only when the dynamics ambiguity set verifies a certain rectangularity property, that amounts to an independence assumption between the uncertainty sets of each transition and is related to time-varying uncertainty.
They do not consider continuous states, nor handle time-invariant uncertainty. We do so by not relying on a dynamic programming evaluation of the robust cost-to-go, but on monte-carlo estimates along a full trajectory.

@@

We also assume to know an upper-bound $S$ over $\|\theta\|_\infty$.
\end{assumption}

@@

\begin{theorem}[Confidence ellipsoid of \citep{Abbasi2011}, Matrix version]
\label{thm:confidence_ellipsoid}
Under \autoref{assumpt:gaussian-noise}, it holds with probability at least $1-\delta$ that
\begin{align}
    \label{eq:confidence-ellipsoid}
    \| \theta_{N,\lambda}  - \theta\|_{G_{N,\lambda}} \leq \beta_N(\delta),
\end{align}
with
\begin{equation}
    \beta_N(\delta)\eqdef \sqrt{2\ln \left(\frac{\det(G_{N,\lambda})^{1/2}}{\delta\det(\lambda I_d)^{1/2}}\right)}
     + (\lambda d)^{1/2}S.
\end{equation}
\end{theorem}

@@

Another strategy presented in \autoref{sec:tight-polytope} produces a much tighter polytope, at the price of an increased computational cost required by the diagonalisation of $G_{N,\lambda}$.

@@

However, \citet{leurent2019interval} showed that this predictor can have unstable dynamics, even for stable systems, which can cause a rapid build-up of uncertainty and produce over-conservative behaviours. They propose an enhanced predictor which produces tighter and more stable predictions, and whose stability can be assessed by solving a linear matrix inequality, at the price of an additional requirement:

\begin{assumption}
\label{assumpt:metzler}
There exists an orthogonal matrix $P\in\Real^{d\times d}$ such that $P^\transp A_0 P$ is Metzler, that is, all non-diagonal coefficients are non-negative.
\end{assumption}
In practice, this assumption is often verified. It is for instance the case whenever $A_0$ is diagonalisable, and a fortiori when it is stabilisable. The similarity transformation of \citep{Efimov_a2013} provides a method to compute such $P$ whenever there exist $C_1,C_2$ which make $A(\theta, C_1)$ and $(A_0, C_2)$ observable.
Under this assumption, without loss of generality up to a change of basis $x\rightarrow P^{-1}x$ that we do not write for the sake of readability, this assumption is equivalent to considering that $A_0$ itself is Metzler.

@@

We start by showing a preliminary proposition:

Let $\{\Phi_n\}_{n=1}^\infty$ be an $\Real^{p\times d}$-valued stochastic process such that $\Phi_n$ is $F_n$-measurable. Assume that $G$ is a $d\times d$ positive definite matrix. For any $n\geq 0$, define:
\begin{equation*}
    \overline{G}_n = G + \sum_{s=1}^n \Phi_s^\transp \Sigma_p^{-1} \Phi_s \in \Real^{d\times d} \quad S_n = \sum_{s=1}^n \Phi_s^\transp\Sigma_p^{-1}\eta_s \in \Real^{d}.
\end{equation*}
Then, for any $\delta>0$, with probability at least $1-\delta$, for all $n\geq0$,
\begin{align*}
\| S_n \|_{\overline{G}_n^{-1}} \geq \sqrt{2\ln \left(\frac{\det\left(\overline{G}_n\right)^{1/2}}{\delta\det(G)^{1/2}}\right)}.
\end{align*}
\end{proposition}
\begin{proof}
Let 
\begin{equation*}
    G_t = \sum_{s=1}^t \Phi_s^\transp \Sigma_p^{-1} \Phi_s \in \Real^{d\times d}
\end{equation*}
And for any $z\in\Real^d$,
\begin{equation*}
    M_t^z = \exp{\left(\inp{z}{S_t} - \frac{1}{2}\|z\|_{G_t}\right)}
\end{equation*}
\begin{equation*}
    D_t^z = \exp{\left(\inp{\Phi_t z}{\eta_t}_{\Sigma_p^{-1}} - \frac{1}{2}\|\Phi_t z\|_{\Sigma_p^{-1}}\right)}
\end{equation*}
Then,
\begin{align*}
    M_t^z &= \exp{\left(\sum_{s=1}^t z^\transp \Phi_s^\transp \Sigma_p^{-1} \eta_s - \frac{1}{2} (\Phi_s z)^\transp\Sigma_p^{-1}(\Phi_s z) \right)} \\
    &= \prod_{s=1}^{t} D_s^z
\end{align*}
and using the Sub-Gaussianity of $\eta_t$
\begin{align*}
    \expectedvalue\left[D_t^z \condbar F_{t-1}\right] = {}& \exp{\left(- \frac{1}{2}\|\Phi_t z\|_{\Sigma_p^{-1}}\right)}\\ &\expectedvalue\left[\exp{\left(\inp{\Phi_t z}{\eta_t}_{\Sigma_p^{-1}}\right)} \condbar F_{t-1}\right]  \\
    \leq {} & \exp{\left(- \frac{1}{2}\|\Phi_t z\|_{\Sigma_p^{-1}}\right)}\\
    &\exp{\left((z^\transp \Phi_t^\transp \Sigma_p^{-1})\Sigma_p(\Sigma_p^{-1} \Phi_t z)\right)}\\
    &= 1
\end{align*}
\begin{align*}
    \expectedvalue\left[M_t^z \condbar F_{t-1}\right] = \left(\prod_{s=1}^{t-1} D_s^z\right) \expectedvalue\left[D_t^z \condbar F_{t-1}\right] \leq M_{t-1}^z
\end{align*}
Showing that $(M_t^z)_{t=1}^\infty$ is indeed a supermartingale and in fact $\expectedvalue[M_t^z]\leq 1$.
It then follows by Doob's upcrossing lemma for supermartingale that $M_\infty^z = \lim_{t\to\infty} M_t^z$ is almost surely well-defined, and so is $M_\tau^z$ for any random stopping time $\tau$.

@@

@@

Having shown this preliminary result, we move on to the proof of \autoref{thm:confidence_ellipsoid}.

\begin{proof}
For all $x\in\Real^d$, \eqref{eq:vector_rls} gives:
\begin{align*}
    x^\transp\theta_{N,\lambda}  -x^\transp\theta &= x^\transp G_{N, \lambda}^{-1}\sum_{n=1}^N \Phi_n^\transp \Sigma_p^{-1}\eta_n
    - \lambda x^\transp G_{N, \lambda}^{-1}\theta\\
    &= \inp{x}{\sum_{n=1}^N \Phi_n^\transp \Sigma_p^{-1}\eta_n}_{G_{N, \lambda}^{-1}} - \lambda\inp{x}{\theta}_{G_{N, \lambda}^{-1}}
\end{align*}
